
---
title: "Distance and dissimilarities"
output:
  html_notebook:
    toc: yes
  html_document:
    toc: yes
    df_print: paged
  pdf_document:
    toc: yes
---




```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```


# Definition of a distance

* A distance function or a metric on $\mathbb{R}^n,\:n\geq 1$, is a function $d:\mathbb{R}^n\times\mathbb{R}^n\rightarrow \mathbb{R}$.
* A distance function must satisfy some required properties or axioms. 
* There are three main axioms.
* A1. $d(\mathbf{x},\mathbf{y})= 0\iff \mathbf{x}=\mathbf{y}$ (identity of indiscernibles);
* A2. $d(\mathbf{x},\mathbf{y})= d(\mathbf{y},\mathbf{x})$ (symmetry);

* A3. $d(\mathbf{x},\mathbf{z})\leq d(\mathbf{x},\mathbf{y})+d(\mathbf{y},\mathbf{z})$  (triangle inequality),
where $\mathbf{x}=(x_1,\cdots,x_n)$, $\mathbf{y}=(y_1,\cdots,y_n)$ and $\mathbf{z}=(z_1,\cdots,z_n)$ are all vectors of $\mathbb{R}^n$.
* We should use the term _dissimilarity_ rather than _distance_ when not all the three axioms A1-A3 are valid.
* Most of the time, we shall use, with some abuse of vocabulary, the term distance.

# Exercice 1

* Prove that the three axioms A1-A3 imply the non-negativity condition: $$d(\mathbf{x},\mathbf{y})\geq 0.$$


# Euclidean distance


* It is defined by:

$$d(\mathbf{x},\mathbf{y})=\sqrt{\sum_{i=1}^n (x_i-y_i)^2}.$$
* A1A2 ae onbvious.
* The proof of A3 is provided below.


# Manhattan distance

* The Manhattan distance also called  taxi-cab metric or city-block metric is defined by:

$$d(\mathbf{x},\mathbf{y})
=\sum_{i=1}^n |x_i-y_i|.
$$

* A1-A2 hold.
* A3 also holds using the fact that $|a+b|\leq |a|+|b|$ for any reals $a,b$.
* There exists also a weighted version  of the Manhattan distance called the Canberra distance.

[Manhattan distance vs Euclidean distance Graph](https://upload.wikimedia.org/wikipedia/commons/0/08/Manhattan_distance.svg)

```{r}
x = c(0, 0)
y = c(6,6)
dist(rbind(x, y), method = "euclidian")
6*sqrt(2)
dist(rbind(x, y), method = "manhattan")
```


# Canberra distance

* It is defined by:

$$d(\mathbf{x},\mathbf{y})
=\sum_{i=1}^n \frac{|x_i-y_i|}{|x_i|+|y_i|}.$$

* Note that the term $|x_i − y_i|/(|x_i|+|y_i|)$ is not properly defined when $x_i=y_i=0$.
* By convention we set the ratio to be zero in that case.
* The Canberra distance is specially sensitive to small changes near zero.


```{r}
x = c(0, 0)
y = c(6,6)
dist(rbind(x, y), method = "canberra")
6/6+6/6
```


# Exercice 2

* Prove that the Canberra distance is a true distance.

# Minkowski distance
* Both the Euclidian and the Manattan distances are special cases of  the Minkowski distance which is defined, for $p\geq 1$, by: 
$$
d(\mathbf{x},\mathbf{y})=
\left[\sum_{i=1} |x_i-y_i|^{p}\right]^{1/p}.
$$
 * For $p=1$, we get the Manhattan distance.
 * For $p=2$, we get the Euclidian distance.
* Let us also define: 
$$\|\mathbf{x}\|_p\equiv\left[\sum_{i=1}^n |x_i|^{p}\right]^{1/p},$$
where $\|\mathbf{\cdot}\|_p$ is known as the $p$-norm or Minkowski norm.
* Note that the Minkowski distance and norm are related by:
$$
d(\mathbf{x},\mathbf{y})=\|\mathbf{x}-\mathbf{y}\|_p.
$$
* Conversely, we have:
$$
\|\mathbf{x}\|_p=d(\mathbf{x},\mathbf{0}),
$$
where $\mathbf{0}$ is the null-vetor of $\mathbb{R}^n$.

```{r}
library("ggplot2")
x = c(0, 0)
y = c(6,6)
MinkowDist=c()
for (p in seq(1,30,.01))
{
MinkowDist=c(MinkowDist,dist(rbind(x, y), method = "minkowski", p = p))     
}
ggplot(data =data.frame(x = seq(1,30,.01), y=MinkowDist ) , mapping = aes(x = x, y = y))+geom_point(size=.1,color="red")+xlim(1,11)+xlab("p")+ylab("Minkowski Distance")+ggtitle("Minkowski distance wrt p")
```

# Chebyshev distance 
* At the limit, we get the Chebyshev distance which is defined by:
$$
d(\mathbf{x},\mathbf{y})=\max_{i=1,\cdots,n}(|x_i-y_i|)=\lim_{p\rightarrow\infty}
\left[\sum_{i=1} |x_i-y_i|^{p}\right]^{1/p}.
$$
* The corresponding norm is:
$$
\|\mathbf{x}|_\infty=\max_{i=1,\cdots,n}(|x_i|).
$$

# Minkowski inequality

* The proof of the triangular inequality A3 is based on the Minkowski inequality:
* For any nonnegative real numbers $a_1,\cdots,a_n$; $b_1,\cdots,b_n$, and for any $p\geq 1$, we have:
$$
\left[\sum_{i=1}^n (a_i+b_i)^{p}\right]^{1/p}\leq
\left[\sum_{i=1}^n a_i^{p}\right]^{1/p}
+\left[\sum_{i=1}^n b_i^{p}\right]^{1/p}.
$$
* To prove that the Minkowski distance satisfies A3, notice that 
$$
 \sum_{i=1}^n|x_i-z_i|^{p}= \sum_{i=1}^n|(x_i-y_i)+(y_i-z_i)|^{p}.
$$
* Since for any reals $x,y$, we have: $|x+y|\leq |x|+|y|$, and using the fact that $x^p$ is increasing in $x\geq 0$, we obtain:
$$
 \sum_{i=1}^n|x_i-z_i|^{p}\leq \sum_{i=1}^n(|x_i-y_i|+|y_i-z_i|)^{p}.
$$

* Applying the Minkowski inequality with $a_i=|x_i-y_i|$ and $b_i=|y_i-z_i|$, $i=1,\cdots,n$, we get:
$$
 \sum_{i=1}^n|x_i-z_i|^{p}\leq \left(\sum_{i=1}^n |x_i-y_i|^{p}\right)^{1/p}+\left(\sum_{i=1}^n |y_i-z_i|^{p}\right)^{1/p}.
$$

# Hölder inequality

* The proof of the Minkowski inequality itself requires the Hölder inequality:
* For any nonnegative real numbers $a_1,\cdots,a_n$; $b_1,\cdots,b_n$, and any $p,q>1$ with $1/p+1/q=1$, we have:
$$
\sum_{i=1}^n a_ib_i\leq
\left[\sum_{i=1}^n a_i^{p}\right]^{1/p}
\left[\sum_{i=1}^n b_i^{q}\right]^{1/q}
$$
* The proof of the Hölder inequality relies on the Young inequality:
* For any $a,b>0$, we have
$$
ab\leq \frac{a^p}{p}+\frac{b^q}{q},
$$
with equality occuring iff: $a^p=b^q$. 
* To prove the Young inequality, one can use the (strict) convexity of the exponential function.
* For any reals $x,y$, we have:
$$
e^{\frac{x}{p}+\frac{y}{q} }\leq \frac{e^{x}}{p}+\frac{e^{y}}{q}. 
$$
* We then set: $x=p\ln a$ and $y=q\ln b$ to get the Young inequality.
* A good reference on inequalities is: Z. Cvetkovski,  Inequalities: theorems, techniques and selected problems, 2012, Springer Science & Business Media.
 # Cauchy-Schwartz inequality
* Note that the triangular inequality for the Minkowski distance implies: 
$$
\sum_{i=1}^n |x_i|\leq
\left[\sum_{i=1}^n |x_i|^{p}\right]^{1/p}.
$$
* Note that for $p=2$, we have $q=2$. The Hölder inequality implies for that special case
$$
\sum_{i=1}^n|x_iy_i|\leq\sqrt{\sum_{i=1}^n x_i^2}\sqrt{\sum_{i=1}^n y_i^2}. 
$$
* Since the LHS od thes above inequality is greater then $|\sum_{i=1}^nx_iy_i|$, we get the Cauchy-Schwartz inequality

$$
|\sum_{i=1}^nx_iy_i|\leq\sqrt{\sum_{i=1}^n x_i^2}\sqrt{\sum_{i=1}^n y_i^2}. 
$$
* Using the dot product notation called also scalar product noation: $\mathbf{x\cdot y}=\sum_{i=1}^nx_iy_i$, and the norm notation $\|\mathbf{\cdot}\|_2 \|$, the Cauchy-Schwart inequality is:
$$
|\mathbf{x\cdot y} | \leq \|\mathbf{x}\|_2 \| \mathbf{y}\|_2.
$$

# Pearson correlation distance 

* The Pearson correlation coefficient is a similarity measure on $\mathbb{R}^n$ defined by:
$$
\rho(\mathbf{x},\mathbf{y})=
\frac{\sum_{i=1}^n (x_i-\bar{\mathbf{x}})(y_i-\bar{\mathbf{y}})}{{\sqrt{\sum_{i=1}^n (x_i-\bar{\mathbf{x}})^2\sum_{i=1}^n (y_i-\bar{\mathbf{y}})^2}}},
$$
where $\bar{\mathbf{x}}$ is the mean of the vector $\mathbf{x}$ defined by: 
$$\bar{\mathbf{x}}=\frac{1}{n}\sum_{i=1}^n x_i,$$
* Note that the Pearson correlation coefficient satisfies P2 and  is invariant to any positive linear transformation, i.e.: $$\rho(\alpha\mathbf{x},\mathbf{y})=\rho(\mathbf{x},\mathbf{y}),$$ for any $\alpha>0$.
* The Pearson distance (or correlation distance) is defined by:
$$
d(\mathbf{x},\mathbf{y})=1-\rho(\mathbf{x},\mathbf{y}).$$

* Note that the Pearson distance does not satisfy A1 since $d(\mathbf{x},\mathbf{x})=0$ for any non-zero vector $\mathbf{x}$. It neither satisfies the triangle inequality. However, the symmetry property is fullfilled. 

# Cosine correlation distance

* The cosine of the angle $\theta$ between two vectors $\mathbf{x}$ and $\mathbf{y}$ is a measure of similarity given by:
$$
\cos(\theta)=\frac{\mathbf{x}\cdot \mathbf{y}}{\|\mathbf{x}\|_2\|\mathbf{y}\|_2}=\frac{\sum_{i=1}^n x_i y_i}{{\sqrt{\sum_{i=1}^n x_i^2\sum_{i=1}^n y_i^2}}}.
$$
* Note that the cosine of the angle between the two centred vectors $\mathbf{x}-\bar{\mathbf{x}}\mathbf{1}$ and $\mathbf{y}-\bar{\mathbf{y}}\mathbf{1}$ coincides with the Pearson correlation coefficient of $\mathbf{x}$ and $\mathbf{y}$, where $\mathbf{1}$ is a vector of units of $\mathbb{R}^n$.
* The cosine correlation distance is defined by:
$$
d(\mathbf{x},\mathbf{y})=1-\cos(\theta).
$$
* It shares similar properties than the Pearson correlation distance. Likewise, Axioms A1 and A3 are not satisfied.

# Spearman correlation distance 

* To calculate the Spearman's rank-order correlation, we need to map seperately each of the vectors to ranked data values: 
$$\mathbf{x}\rightarrow \text{rank}(\mathbf{x})=(x_1^r,\cdots,x_n^r).$$
* Here, $x_i^r$ is the rank of $x_i$ among the set of values of $\mathbf{x}$.
* We illustrate this transformation with a simple example:
* If $\mathbf{x}=(3, 1, 4, 15, 92)$, then the rank-order vector is $\text{rank}(\mathbf{x})=(2,1,3,4,5)$.  

```{r}
x=c(3, 1, 4, 15, 92)
rank(x)
```

* The Spearman's rank correlation of two numerical variables $\mathbf{x}$  and $\mathbf{y}$ is simply the Pearson correlation of the two correspnding rank-order variables $\text{rank}(\mathbf{x})$ and $\text{rank}(\mathbf{y})$, i.e. $\rho(\text{rank}(\mathbf{x}),\text{rank}(\mathbf{y}))$. This measure is is useful because it is more robust against outliers than the Pearson correlation.
* If all  the $n$  ranks are distinct, it can be computed using the following formula:
$$
\rho(\text{rank}(\mathbf{x}),\text{rank}(\mathbf{y}))=1-\frac{6\sum_{i=1}^n d_i^2}{n(n^2-1)},
$$
where $d_i=x_i^r-y_i^r,\:i=1,\cdots,n$.
 * The spearman distance is then defined by:
$$
d(\mathbf{x},\mathbf{y})=1-\rho(\text{rank}(\mathbf{x}),\text{rank}(\mathbf{y})).
$$
* It can be shown that easaly that it is not a proper distance.
* If all  the $n$  ranks are distinct, we get:
$$
d(\mathbf{x},\mathbf{y})=\frac{6\sum_{i=1}^n d_i^2}{n(n^2-1)}.
$$

```{r}
x=c(3, 1, 4, 15, 92)
rank(x)
y=c(30,2 , 9, 20, 48)
rank(y)
d=rank(x)-rank(y)
d
cor(rank(x),rank(y))
1-6*sum(d^2)/(5*(5^2-1))
```


# Kendall tau distance 

* The Kendall rank correlation coefficient is calculated from the number of correspondances between the rankings of $\mathbf{x}$ and the rankings of $\mathbf{y}$.
*   The number of pairs of observations among $n$ observations or values is: 
$${n \choose 2} =\frac{n(n-1)}{2}.$$
* The pairs of observations $(x_{i},x_{j})$  and  $(y_{i},y_{j})$ are said to be _concordant_ if: $$\text{sign}(x_j-x_i)=\text{sign}(y_j-y_i),$$ and to be _discordant_ if:  $$\text{sign}(x_j-x_i)=-\text{sign}(y_j-y_i),$$
where $\text{sign}(\cdot)$ returns  $1$ for positive numbers and  $-1$ negative numbers and $0$ otherwise.
* If $x_i=x_j$ or $y_i=y_j$ (or both), there is a tie.
* The Kendall $\tau$ coefficient is defined by (neglecting ties):
$$\tau =\frac {1}{n(n-1)}\sum_{i=1}^{n}\sum_{j=1}^n\text{sign}(x_j-x_i)\text{sign}(y_j-y_i).$$
* Let $n_c$ (resp. $n_d$) be the number of concordant (resp. discordant) pairs, we have $$\tau =\frac {2(n_c-n_d)}{n(n-1)}.$$ 
* The Kendall tau distance is then: $$d(\mathbf{x},\mathbf{y})=1-\tau. $$
* Remark: the triangular inequality may fail in cases where there are ties.

```{r}
x=c(3, 1, 4, 15, 92)
y=c(30,2 , 9, 20, 48)
tau=0
for (i in 1:5)
{  
tau=tau+sign(x -x[i])%*%sign(y -y[i])
}
tau=tau/(5*4)
tau
cor(x,y, method="kendall")
```

# Variables standardization

* Variables are often standardized before measuring dissimilarities.
* Standardization converts the original variables into uniteless variables.
* A well known method is the z-score transformation: 
$$
\mathbf{x}\rightarrow (\frac{x_1-\bar{\mathbf{x}}}{s_\mathbf{x}},\cdots,\frac{x_n-\bar{\mathbf{x}}}{s_\mathbf{x}}),
$$ 
where $s_\mathbf{x}$ is the sample standard deviation given by:
$$
s_\mathbf{x}=\frac{1}{n-1}\sum_{i=1}^n(x_i-\bar{\mathbf{x}})^2.
$$
* The transformed variable will have a mean of $0$ and a variance of $1$.
* The result obtained with Pearson correlation measures and standardized Euclidean distances are comparable.
* For other methods, see: Milligan, G. W., & Cooper, M. C. (1988). A study of standardization of variables in cluster analysis. _Journal of classification_, _5_(2), 181-204.

```{r}
x=c(3, 1, 4, 15, 92)
y=c(30,2 , 9, 20, 48)
(x-mean(x))/sd(x)
scale(x)
(y-mean(y))/sd(y)
scale(y)
```

# Similarity measures for binary data
* We consider here the case where all the $n$ variables are binary.
* 
* A large number of similarity measures have been proposed for binary data. 
* All the measures are defined in terms of the entries in a cross-classification of the counts of matches and mismatches in the p variables for two individuals; 
the general version of this cross-classification is shown in Table 3.2.
* A list of some of the similarity measures that have been suggested for binary data is shown in Table 3.3; a more extensive list can be found in Gower and Legendre (1986). 
* The reason for such a large number of possible measures has to do with the apparent uncertainty as to how to deal with the count of zero–zero matches(d in Table 3.2). In some cases, of course, zero–zero matches are completely equivalent to one–one matches, and therefore should be included in the calculated similarity measure. An example is gender, where there is no preference as to which of the two categories should be coded zero or one. But in other cases the inclusion or otherwise of d is more problematic; for example, when the zero category corresponds to the genuine absence of some property, such as wings in a study of insects. The question that then needs to be asked is do the co-absences contain

|              	| Counts of binary outcomes|     	|     	|           	|
|---	|:---	|:---:	|-----	|-----------	|
|||1|
|1|
|0|


# Distance matrix computation
* We’ll use a subset of the data USArrests
*  We’ll use only a by taking 15 random rows among the 50 rows in the data set. 
* Next, we standardize the data using the function scale():

```{r}
install.packages("FactoMineR")
library("FactoMineR")
data("USArrests") # Loading
head(USArrests, 3) # Print the first 3 rows
set.seed(123)
ss <- sample(1:50, 15) # Take 15 random rows
df <- USArrests[ss, ] # Subset the 15 rows
df.scaled <- scale(df) # Standardize the variables
```

<!--stackedit_data:
eyJoaXN0b3J5IjpbLTE2OTY4MDQ4ODcsODUxOTgzMDIyLC0xOD
U4MDc5NzgxLC00NDAwMjM5NSwzMzcwNzQ2NTMsLTI1NzA1ODY2
OSwtMTYzMTYyMDYzMiw0NDIzNzU2NzQsMzg4NjM3OTQ2LDQ5Mz
A1MzQ3NSwtMTUyMTQ5MjExOSw3MDM0MTk2MjUsLTE1MjE0OTIx
MTksLTE1MjE0OTIxMTksMTY4NTM5OTkxXX0=
-->